---
title: "Assignment 1"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      root.dir = 'data', fig.align = 'center',
                      fig.width = 2.5, fig.height=2.5,
                      cache = TRUE)
```



```{r}
library(tidyverse) # essential tools
library(ggdendro)
```


# Question 1

For the dataset given, do these calculations by hand:

  - Calculate a distance matrix using squared euclidean distance.
  - Use hierarchical clustering with single linkage to cluster the data. 
  Draw the dendro- gram and identify the two-cluster solution.
  - Use hierarchical clustering with average linkage to cluster the data. 
  Draw the den- drogram and identify the two-cluster solution.
  - Cluster the data using kmeans with k = 2. Use starting clusters
  of (a,b,c) and (d,e).

```{r}
set.seed(123)
# Generating data
x <- matrix(sample(-5:5, 10), nrow = 5)
rownames(x) <- letters[1:5]
colnames(x) <- c("U", "V")
x

dx <- dist(x)^2 # finding distances
dx

# building  clusters
hs <- hclust(dx, "single")
ha <- hclust(dx, "average")


# ggplot version 
ggdendrogram(as.dendrogram(hs))
ggdendrogram(as.dendrogram(ha))

c1 <- apply(x[1:3,], 2, mean)
c2 <- apply(x[4:5,], 2, mean)


kmeans(x, centers = rbind(c1, c2), algorithm = "Lloyd")

x %>% 
  as.data.frame() %>% 
  ggplot(aes(U, V)) +
  geom_point(colour = "orange", size = 2) +
  theme_bw()
```

# Question 2 

Eight online shoppers buy 8, 11, 7 , 6, 5, 6 , 7, 8 pairs of socks. The
same eight shoppers buy 0, 0, 0, 0, 1, 1, 1, 1 computers. 

# a)  If you run kmeans on this data with $k = 2$, 
with no scaling, what result would you expect?

```{r}
my_df <- data.frame(x = c(8, 11, 7, 6, 5, 6, 7, 8), 
                 y = c(0, 0, 0,0, 1, 1, 1, 1))

my_df %>% 
ggplot(aes(x, factor(y))) +
  geom_point(colour = "orange", size = 2) +
  theme_bw() +
  geom_text(label = 1:8,hjust= -0.7, vjust = 0.3) +
  labs(y = 'y')

kmeans(my_df, 2)$cluster
kmeans(my_df, 2, nstart = 10)$cluster
```

NOTE: use nstart to run the algorithm from 10 random starts. Better convergence.
Point 2 is in a cluster of its own! 


# b) If both variables are scaled to unit standard deviation, what will
kmeans with $k = 2$ give you?

```{r}
d1 <- my_df %>% 
  mutate_all(scale, center = FALSE)

d1 %>% 
ggplot(aes(x, y)) +
  geom_point(colour = "orange", size = 2) +
  theme_bw() +
  geom_text(label = 1:8,hjust= -0.7, vjust = 0.3) +
  labs(y = 'y')

kmeans(d1, 2, nstart = 10)
```
Points 1-4 in 1 cluster, 5-8 in the other.

# c) Suppose socks cost 2 euro and the computer is 2000 euro. What is you 
clustered the amount spent by each customer using kmeans with
$k = 2$, with no scaling?

```{r}
d <- my_df %>% 
  mutate(x = 2*x, y = 2000*y)

d %>% 
ggplot(aes(x, y)) +
  geom_point(colour = "orange", size = 2) +
  theme_bw() +
  geom_text(label = 1:8,hjust= -0.7, vjust = 0.3) +
  labs(y = 'y')

kmeans(d, 2, nstart = 10)
```

Points 1-4 in 1 cluster, 5-8 in the other. 

# Question 3

The file `eupop.txt` contains the population and percentage distribution
by age for EU countries in 1999. The age categories are 0-14 years, 
15-44 years, 45-64 years and 65 years and over.

# a) Construct the euclidean distance matrix of the percentage variables.
Use it to cluster the countries, using average linkage. Draw the dendrogram 
and interpret. Are there any outlier countries?

```{r}
eupop <- read.table("data/eupop.txt") %>% 
  select(-5)
d <- dist(eupop)
h <- hclust(d, "average")
ggdendrogram(as.dendrogram(h))
```

Ireland is an outlier.

# b) Examine the 3-cluster solution. Which countries belong to each of 
the three clusters? Summarise the partitions with sumPartition 
(in h1code.R) Interpret your findings.

```{r}
source('code/h1code.R')
sumPartition(eupop, cutree(h,3))
```

Ireland is in Cluster 3. 
Germany Greece, Italy, Spain are in Cluster 2. 
Everyone else is in Cluster 1.
Cluster 3: highest proportion of children(under 15), lowest percentage of over 65. Cluster 2: below average for under 15s and above average proportion of over 65s. Cluster 2 and 3 are furtherest apart, cluster 1 and 2 are closest. Cluster 2 is more compact than cluster 1. 

# c) Use the kmeans algorithm to find another 3-cluster grouping of 
countries. Which countries belong to each of the three clusters?

```{r}
km <- kmeans(eupop, 3, nstart = 10)
km
```


Cluster agreement with 3 cluster solution of hclust, except, Portugal and 
Austria are clustered with Greece, Italy, Germany and Spain. This cluster 
still has lower proportion of children and above average porportion of over 65s.


# d) Construct a stars plot which shows the data and clustering obtained 
from kmeans. Optional: can you think of a better way of showing the clusters? 
Can you think of a way to present the data and the clustering results of
both methods on the same graphical display?

```{r, fig.width = 3.5, fig.height=3.5}
clusk <- km$cluster 
o <- order(clusk)
stars(eupop[o, ], nrow = 3, col.stars = clusk[o] + 1)

eupop[o,] %>% 
  mutate(country = rownames(.)) %>% 
  gather(key, value, -country) %>% 
  ggplot(aes(y = value, x = country, fill = key)) +
  geom_bar(stat = "identity") +
  labs(x = "percentage") +
  coord_flip() +
  ggpomological::scale_fill_pomological()


library(GGally)
eupop %>% 
  mutate(col = clusk+1) %>% 
  ggpairs(aes(colour = factor(col), alpha = 0.4))

```

\newpage

# Question 4

Analyzing the music data. 

# a) Run the k-means algorithm over the range $k = 1,\dots, 15$
clusters and record the total within cluster sum of squares (TWSS). 
Let nstart = 25. Plot k versus TWSS and choose the best fitting number 
of clusters. What do you observe? Note: remember to scale the data.

```{r}
music <- read.table("data/music.txt")
music_feat <- music %>% 
  select(3:7) %>% mutate_all(scale)

results <- 
  data.frame(
    tot.withinss = 1:30 %>% 
      purrr::map(kmeans, x = music_feat, nstart = 25) %>% 
      purrr::map_dbl("tot.withinss"),
    ind = 1:30)


results %>% 
ggplot(aes(ind, tot.withinss)) +
  geom_line(colour = 'grey') +
  geom_point(colour = "orange", size = 2) +
  theme_bw() +
  labs(y = 'TWSS', x = "Number of clusters")

```

TWSS declines slowly. Data does not partition into few, small, 
well-defined compact clusters

# b) Make a table of artist vs cluster solution from k = 5.

```{r}
clusk <- kmeans(music_feat, centers = 5, nstart = 25)$cluster


music %>% 
  mutate(clusk = clusk) %>% 
  group_by(Artist, clusk) %>% 
  count()
```


All but one Abba tracks in a single cluster. Most of Beatles an the 
Eels tracks in a single cluster.

\newpage
# Question 5

Protein data. We want to study the similarities and differences in the 
protein composition of the diets of different countries. Using any methods 
that you choose from this course or otherwise, write a brief summary.


```{r}
protein <- read.table("data/protein.txt")
protein_feat <- protein %>%  select(2:10)
row.names(protein_feat) <- protein$Country
d <- dist(protein_feat)
h <- hclust(d, "complete")
ggdendrogram(as.dendrogram(h))

hc <- cutree(h, 5)


results <- 
  data.frame(
    tot.withinss = 1:20 %>% 
      purrr::map(kmeans, x = protein_feat, nstart = 10) %>% 
      purrr::map_dbl("tot.withinss"),
    ind = 1:20)

results %>% 
ggplot(aes(ind, tot.withinss)) +
  geom_line(colour = 'grey') +
  geom_point(colour = "orange", size = 2) +
  theme_bw() +
  labs(y = 'TWSS', x = "Number of clusters")


clusk <-  kmeans(protein_feat, centers = 5, nstart = 10)$cluster
o <- order(clusk)

```


```{r, fig.width=5, fig.height=4}
stars(protein_feat[o,], nrow = 3, col.stars = clusk[o]+1)

table(clusk, hc)
```


