---
title: "Assignment 1"
output: pdf_document
author: Name Student no.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Q1

```{r fig.width=3,fig.height=3}
set.seed(123)
x <- matrix(sample(-5:5, 10), nrow=5)
rownames(x)<- letters[1:5]
colnames(x)<- c("U", "V")
x
dx <- dist(x)^2
dx
hs <- hclust(dx, "single")
ha <- hclust(dx, "average")
plot(as.dendrogram(hs))
plot(as.dendrogram(ha))

c1 <- apply(x[1:3,], 2, mean)
c2 <- apply(x[4:5,], 2, mean)

kmeans(x, centers= rbind(c1, c2), algorithm ="Lloyd")
plot(x)

```

# Q2

#a)

```{r}
x <- c(8,11,7,6,5,6,7,8)
y <- c(0,0,0,0,1,1,1,1)
plot(x,y, xlim = c(5, 12), asp=1)
text(x+.3,y, 1:8 )
d <- data.frame(x=x,y=y)
kmeans(d,2)$cluster
kmeans(d,2, nstart=10)$cluster
```

NOTE: use nstart to run the algorithm from 10 random starts. Better convergence.
Point 2 is in a cluster of its own


#b)

```{r}

d1 <- scale(d, center=F)
plot(d1,  xlim = c(0.5,1.5))
text(d1[,1]+0.03,d1[,2], 1:8)
kmeans(d1,2, nstart=10)

```

Points 1-4 in 1 cluster, 5-8 in the other

#c)

```{r}
d <- data.frame(x=2*x,y=2000*y)
plot(d)
text(d[,1]+ 0.5,d[,2], 1:8)
kmeans(d,2, nstart=10)
```

Points 1-4 in 1 cluster, 5-8 in the other

\newpage
# Q3

# a)
```{r}
eupop <- read.table("data/eupop.txt", header=T, row.names=1)
eupop <- eupop[,-5]
d <- dist(eupop)
h<- hclust(d, "average")
plot(as.dendrogram(h))
```
Ireland is outlier.

# b)

```{r}
source('code/h1code.R')
sumPartition(eupop, cutree(h,3))
```

Ireland is in Cluster 3. 
Germany Greece, Italy, Spain are in Cluster 2. 
Everyone else is in Cluster 1.
Cluster 3: highest proportion of children(under 15), lowest percentage of over 65. Cluster 2: below average for under 15s and above average proportion of over 65s. Cluster 2 and 3 are furtherest apart, cluster 1 and 2 are closest. Cluster 2 is more compact than cluster 1. 

#c)
```{r}
km <- kmeans(eupop, 3,nstart=10)
km
```


Cluster agreement with 3 cluster solution of hclust, except, Portugal and Austria are clustered with Greece, Italy, Germany and Spain. This cluster still has lower proportion of children and above average porportion of over 65s.


#d)

```{r}

clusk <- km$cluster
o <- order(clusk)
stars(eupop[o,],nrow=3, col.stars=clusk[o]+1)

par(mar=c(3,6,3,2))
barplot(t(as.matrix(eupop[o,])), col=1:4, horiz=T, las=2)

 # another display, kmeans is colours, hclust is symbols
pairs(eupop, col = clusk+1, pch = cutree(h,3))


```

\newpage
# Q4

# a)
```{r}
music <- read.csv("data/music.csv")
music.feat <- music[, 4:8]
music.feat <- scale(music.feat)
wss <- vector(length=30)

for (k in 1:30) wss[k] <- kmeans(music.feat,centers=k, nstart = 25)$tot.withinss
plot(1:30, wss, type="b", xlab="Number of Clusters k",
     ylab="TWSS")
```

# b)
TWSS declines slowly. Data does not partition into few, small, well-defined compact clusters. 
```{r}
clusk <- kmeans(music.feat,centers=5, nstart = 25)$cluster

table(music$Artist, clusk)


```

Students may not have set.seed, could have different output due to random starts. 

All but one Abba tracks in a single cluster. Most of Beatles an the Eels tracks in a single cluster.

\newpage
# Q5


Anything sensible.
```{r}
protein <- read.csv("data/protein.csv")
protein.feat <- protein[, 2:10]
row.names(protein.feat) <- protein$Country
d <- dist(protein.feat)
h <- hclust(d, "complete")
plot(as.dendrogram(h))
hc <- cutree(h, 5)


wss <- vector(length=20)
for (k in 1:20) wss[k] <- kmeans(protein.feat,centers=k, nstart = 10)$tot.withinss
plot(1:20, wss, type="b", xlab="Number of Clusters k",
     ylab="TWSS")

clusk <-  kmeans(protein.feat,centers=5, nstart = 10)$cluster
o <- order(clusk)
stars(protein.feat[o,],nrow=3, col.stars=clusk[o]+1)

table(clusk, hc)
```

