---
title: "Assignment 2"
output: pdf_document
author: Name Student no.
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Q1

# a) - e)
```{r fig.width=3,fig.height=3}
x <- matrix(c(4 ,1,-1,-3,1, 2,0,-1,5,-1), nrow=5)
plot(x)

var(x)
cor(x)



xs<- scale(x)

xs
t(xs)%*%(xs)/(5-1)
var(xs)

eigen(cor(x))



prcomp(x, scale=TRUE)


```


\newpage
# Q2

#a)

```{r}

bfat <- read.table("data/bodyfat.txt", header=T)
bfat <- bfat[,c("neck","abdomen", "knee", "ankle")]
pairs(bfat)

bfat[bfat[,4]>30, ]

```

There are two outliers with extreme ankle values, but non 
extreme values on other variables.
They are observations 31 and 84. 

<!-- Observations 35, 40 has high weight. -->
<!-- Observation 36 has high body fat. -->



#b)

```{r}

screeplot <- function(p) {
  e <- p$sdev ^ 2
  e <- e / sum(e)
  plot(
    1:length(e),
    e,
    xlab = "Component number",
    pch = 20,
    ylab = "Variance proportion",
    main = "Scree plot",
    axes = F,
    ylim = c(0, max(e)*1.04)
  )
  lines(1:length(e), e)
  axis(1, at = 1:length(e))
  axis(2)
}

# solution
p <- prcomp(bfat, scale=TRUE)
p$rotation[,1:2]
summary(p)

screeplot(p)


```

66\% of variability explaide by the 1st PC, 85\% by the first 2 PCs and 94\% by the first 3 PCs. 

#c)

```{r}
biplot(p, scale=0, cex=c(.5,.5), cex.axis=.5)
```


The first component is a weighted average of the variables.
It is an overall measure of size.
The second component is a contrast of neck and abdomen with ankle.
It is a measure of the difference between top size and ankle.
The visible outliers are 84 and 31, with the big ankle values.


#d)

```{r}
p<- prcomp(bfat[-c(31,84),], scale=TRUE)
p$rotation[,1:2]
summary(p)

screeplot(p)
biplot(p, scale=0, cex=c(.5,.5), cex.axis=.5)

```

The first component is a weighted average of the variables.
It is an overall measure of size.
The second component is a contrast of neck and abdomen with knee and ankle.
It is a measure of the difference between top size and lower size.
The  high weight people stick out on the first component,
but are not that extreme.

\newpage
# Q3

```{r}



# read in the correlation data as a vector

crimcorr <- matrix(c(
  1.000, 0.402, 0.396, 0.301, 0.305, 0.339, 0.340,
  0.402, 1.000, 0.618, 0.150, 0.135, 0.206, 0.183,
  0.396, 0.618, 1.000, 0.321, 0.289, 0.363, 0.345,
  0.301, 0.150, 0.321, 1.000, 0.846, 0.759, 0.661,
  0.305, 0.135, 0.289, 0.846, 1.000, 0.797, 0.800,
  0.339, 0.206, 0.363, 0.759, 0.797, 1.000, 0.736,
  0.340, 0.183, 0.345, 0.661, 0.800, 0.736, 1.000), nrow = 7, byrow = TRUE)

colnames(crimcorr)<- c("Head-L","Head-B","Face-B",
                     "L-Fing","L-Fore","L-Foot",
                     "Height")
V <- eigen(crimcorr)

V$values/sum(V$values)

# can hack it to use the screeplot above, else make your own.
V$sdev <- sqrt(V$values)
screeplot(V)

```
Proportion variance explained by the 1st PC is 0.54, first two is 0.76 etc.

First PC is a measure of overall size of the person. Second PC contrasts head measurements with the rest. Third PC is the head length etc.

# Q4

# a)

Regression, inference,n = 500, 1 response, 4 predictors.
All predictors are quantitative
except country which is categorical.
Inflexible  better for inference.

# b)
 
classification, prediction,n = 20, 1 response (binary), 12 predictors.
All described predictors are quantitative
Inflexible  better because so many predictors relative to n.


# c)

Regression, inference,n = unknown, 1 response (quantitative),
2 predictors, birthweight quantitative and gender categorical.Inference.
Inflexible  better because to understand predictors response association

# d)

Classification, prediction,n = 32, 1 response (categorical, 3 classes),
56 predictors, structure unknown.
Inflexible  becuase n so large relative to p.
 
 
\newpage
# Q5

# a)
```{r}
# install.packages("ISLR") #home computer, first time only
library(ISLR)
Auto <-Auto[complete.cases(Auto[,c(1,4,5)]),] # to remove NAs
plot(mpg ~ weight, data=Auto)
plot(mpg ~ horsepower, data=Auto)
```

mpg goes down as weight goes up, plot shows curvature.
mpg goes down as hp goes up, plot shows curvature.

# b)

```{r}


library(plot3D) # install package
scatter3D(Auto$weight,Auto$horsepower,Auto$mpg)

library(plot3Drgl)
scatter3Drgl(Auto$weight,Auto$horsepower,Auto$mpg)

```


plot shows that points lie on a surface, not a plane, so a linear fit is not appropriate





# c)


```{r}
set.seed(123)
train <- sample(nrow(Auto), round(.8*nrow(Auto)))
AutoTrain <- Auto[train,]
AutoTest <- Auto[-train,]
```

```{r}
f1 <- lm(mpg~weight+horsepower, data=AutoTrain)
summary(f1)
```
 Both predictors are significant as p values are so small
# d)

```{r}

wt1 <- seq(1610 ,5140, length.out = 30)
hp1 <- seq(45, 230, length.out = 30)
pred <- predict(f1, expand.grid(weight=wt1, horsepower=hp1))
pred <- matrix(pred,30,30)
library(plot3D)
scatter3D(AutoTrain$weight,AutoTrain$horsepower,AutoTrain$mpg, pch = 18, surf = list(x = wt1, y = hp1, z = pred))

```


You can see for high values of z=mpg, points lie far away and
mostly above from the fitted planes
so the linear fit does not look appropriate.

#e)

```{r}
f2 <- loess(mpg~weight+horsepower, data=AutoTrain)
pred <- predict(f2, expand.grid(weight=wt1, horsepower=hp1))
pred <- matrix(pred,30,30)

scatter3D(AutoTrain$weight,AutoTrain$horsepower,AutoTrain$mpg, pch = 18, 
      surf = list(x = wt1, y = hp1, z = pred))

```
It looks to capture the pattern of the association, but a
smoother surface might be better.


#f)

```{r}
mean(residuals(f1)^2)
mean(residuals(f2)^2)
```


the train mse is smaller for f2, so the fit is closer to the observed data

#g)

```{r}
pred1 <- predict(f1, AutoTest)
mean((pred1 - AutoTest$mpg)^2)
pred2 <- predict(f2, AutoTest)
mean((pred2 - AutoTest$mpg)^2, na.rm=T)
```

the test MSE is about the same for both fits. Choose simpler model (f1)
